---
title: "MATH-472: Homework 3"
author: "Andrew Moore"
date: "3/16/23"
knitr:
  opts_chunk: 
    echo: true
    warning: false
    message: false
format:
  html:
    include-in-header:
      text: |
        $$
        \require{cancel}
        \require{physics}
        $$
  pdf:
    include-in-header:
      text: |
        \usepackage{cancel}
        \usepackage{physics}
---

```{r}
#| echo: false

set.seed(20230312)
```

# Question 1

Do 6.1, 6.3, 6.6, 6.9, 6.10 in the exercises of Chapter 6.

## 6.1

Analytically:
$$
\int_0^{\frac{\pi}{3}} sin(t)\ dt = -cos(t) \Bigg|_{t = 0}^{t = \frac{\pi}{3}} = cos(0) - cos(\frac{\pi}{3}) = \frac{1}{2}
$$

Monte-Carlo integration:
```{r}
g <- function(t) sin(t)
u <- runif(10000, 0, pi/3)
mean(g(u))
```

## 6.3

```{r}
g <- function(x) exp(-x)
f <- function(x) sqrt(x)

gen_thetas <- function(N = 1000, n = 200) {
  t <- numeric(N)
  tstar <- numeric(N)
  
  for (i in 1:N) {
    u <- runif(n, 0, 0.5)
    e <- rexp(n, rate = 1)
    t[i] <- mean(g(u))
    tstar[i] <- mean(e <= 0.5)
  }

  c(theta = var(t), theta_star = var(tstar))
}

gen_thetas()
```

## 6.6

```{r}
mc <- function(f, n, r, anti = TRUE) {
  t <- numeric(n)
  for (i in 1:n) {
    m <- runif(r / 2)
    n <- if (anti) 1 - m else runif(r / 2)
    u <- c(m, n)
    t[i] <- mean(f(u))
  }
  return(t)
}

f <- function(x) exp(x)

smc <- mc(f, 1000, 100, anti = FALSE)
ava <- mc(f, 1000, 100)
```

## 6.9

The Rayleigh Density is
$$
f(x) = \frac{x}{\sigma^2} e^{\frac{-x^2}{2\sigma^2}} \text{ where } x \geq 0, \sigma > 0.
$$

Implement a function to generate samples from a $Rayleigh(\sigma)$ distribution using antithetic variables. What is the percent reduction in variance of 
$\frac{X + X'}{2}$ compared with $\frac{X_1 + X_2}{2}$ for independent $X_1$ and $X_2$?

## 6.10

Use Monte Carlo integration with antithetic variables to estimate
$$
\int_0^1 \frac{e^{-x}}{1 + x^2}\ dx,
$$
and find the approximate reduction in variance as a percentage of the variance without variance reduction.

\newpage
# Question 2

Suppose you use the importance sampling method to obtain a Monte Carlo estimate of 
$$
\theta = \int_0^\infty g(x)\ dx,
$$
where
$$
g(x) = \frac{x^2}{\sqrt{2 \pi}} e^{-x^2 / 2}.
$$

(a) A possible importance function for the purpose could be
$$
f(x) = \frac{1}{\Gamma(3/2)} 2^{3/2} x^{3/2-1} e^{-2x}, 1 < x < \infty.
$$
Note that $t = x - 1$ has a gamma distribution with shape 3/2 and rate 2. Draw two functions $y = g(x)$ and $y = f(x)$ on the xy-plane for the following values: `x <- seq(1, 10, 0.01)`.

(b) Estimate $\theta$ using the importance function in (a).

```{r}
g <- function(x) x^2 / sqrt(2 * pi) * exp(-x^2 / 2)
f <- function(x) 1 / gamma(3 / 2) * 2^(3/2) * x^(3) * exp(-2 * x)
x <- seq(1, 10, 0.01)
d <- data.frame(x = x, y0 = g(x), y1 = f(x))

library(ggplot2)

ggplot(d, aes(x = x)) +
  geom_line(aes(y = y0), size = 1.1, color = "grey") +
  geom_line(aes(y = y1), size = 1.1, color = "orange") +
  annotate("text", x = 2.0, y = 0.3, label = "g(x)", color = "grey", size = 4.5) +
  annotate("text", x = 2.5, y = 0.45, label = "f(x)", color = "orange", size = 4.5) +
  theme_minimal(base_size = 14) + 
  labs(x = "x", y = "y")
```